D3064R0: Bringing the Laws of Physics, Computer-System Constraints, and Compilers to Bear on the OOTA Problem

### 2024-01-27 ootavacuous.tex
January 27 application of Alan Stern feedback, #1
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@

Roughly speaking and all other things being equal, \emph{there is semantic
dependency from a given load to a given store when a change in the value
loaded can change the value stored.}

This paper will follow P2055R0~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}
in using the term \emph{full C++} to denote the standard including the
above prohibition, and the term \emph{strict C++} to denote a hypothetical
standard that excludes this prohibition, but is otherwise identical
to full C++.
Unqualified \emph{C++} means full C++.

@@

This paper will use the \co{herd7} tool to analyze fragments of C++ code.
This tool carries out the moral equivalent of full state-space searches
of concurrent code fragments, and further will in some cases identify
OOTA cycles.

@@

\subsection{Semantic Dependencies and Source Code}
\label{sec:Semantic Dependencies and Source Code}

@@

\subsection{Semantic Dependencies Can Be Many To One}
\label{sec:Semantic Dependencies Can Be Many To One}

@@

\subsection{Semantic Dependencies Affected by Optimizations}
\label{sec:Semantic Dependencies Affected by Optimizations}

@@

There are semantic dependencies from \co{y} to \co{x} and from \co{z}
to \co{x}.
However, if the implementation knows that the value of \co{y} is
always the negative of \co{z} there is no semantic dependency and the
implementation can again act as if later accesses to \co{x} were instead
the constant zero, again, at least until the next assignment to \co{x}.
Please note that this leaves aside the question of how the compiler
would know this if \co{y} and \co{z} are atomic objects that cannot be
updated pairwise atomically and are subject to change at any time by
other threads.

\subsection{Semantic Dependencies Affected by \co{if} Statements}
\label{sec:Semantic Dependencies Affected by if Statements}

@@

Because the stores executed on each branch of that \co{if} statement store
identical values to identical addresses, one could equally well regard
the two statements as performing two different stores or as performing
(for all intents and purposes) a single store.
Although software projects relying on ordering from load-to-store control
dependencies and using a variety of compilers should take the single-store
viewpoint~\cite{Howells2009membartxt},
the fact remains that these two viewpoints lead to differing conclusions
about whether this code represents one execution or two.
It is the compiler's choice.

>>>>

\section{Introduction}
\label{sec:Introduction}

@@@ Add brief description and roadmap of paper, including the simplifications
and restrictions.

@@

Roughly speaking, \emph{there is semantic dependency from a given load
to a given store when, all other things being equal a change in the
value loaded can change the value stored.}

This paper will follow P2055R0~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}
in using the term \emph{full C++} to denote the standard including the
above prohibition, and the term \emph{strict C++}\footnote{
	@@@ Future bikeshedding of name.}
% @@@ Alternatives thus far: OOTA-prone C++, OOTA-enabled C++,
% @@@ relaxed C++, OOTA C++, OOTA-agnostic C++.
to denote a hypothetical standard that excludes this prohibition, but
is otherwise identical to full C++.
Unqualified \emph{C++} means full C++.

@@

This paper will use the \co{herd7}\footnote{
	Available at \path{https://github.com/herd/herdtools7}.}
tool to analyze fragments of C++ code.
This tool carries out the moral equivalent of full state-space searches
of concurrent code fragments.
In some cases its output will include executions with OOTA cycles,
on occasion reporting undefined values for the variables involved in
the cycle.

@@

\subsection{Properties of Semantic Dependencies}
\label{sec:Properties of Semantic Dependencies}

\subsubsection{Semantic Dependencies and Source Code}
\label{sec:Semantic Dependencies and Source Code}

@@

\subsubsection{Semantic Dependencies Can Be Many To One}
\label{sec:Semantic Dependencies Can Be Many To One}

@@

\subsubsection{Semantic Dependencies Affected by Optimizations}
\label{sec:Semantic Dependencies Affected by Optimizations}

@@

There are semantic dependencies from \co{y} to \co{x} and from \co{z}
to \co{x}.
However, if the implementation knows that the value of \co{y} is
always the negative of \co{z} there is no semantic dependency and the
implementation can again act as if later accesses to \co{x} were instead
the constant zero, again, at least until the next assignment to \co{x}.
Please note that this leaves aside the question of how the compiler
would know this if \co{y} and \co{z} are atomic objects that cannot be
updated pairwise atomically and are subject to change at any time by
other threads (a point we will return to).

\subsubsection{Semantic Dependencies Affected by \co{if} Statements}
\label{sec:Semantic Dependencies Affected by if Statements}

@@

Because the stores executed on each branch of that \co{if} statement store
identical values to identical addresses, one could equally well regard
the two statements as performing two different stores or as performing
(for all intents and purposes) a single store.
Although software projects relying on ordering from load-to-store control
dependencies and using a variety of compilers should should err on the
side of caution by assuming there is no semantic
dependency~\cite{Howells2009membartxt},
the fact remains that reasonable C++ implementations might disagree as
to whether this code represents one execution or two.
It is the implementations's choice.
